---
title: "Telecom Customer Churn — Predictive Modeling & Retention Strategy"
author: "Adarsh Shukla"
date: "2025"
output:
  html_document:
    toc: true
    toc_float: true
    toc_depth: 3
    theme: flatly
    highlight: tango
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE)
library(tidyverse)
library(caret)
library(ggplot2)
library(pROC)
library(e1071)
library(MASS)
library(naivebayes)
library(rpart)
library(rpart.plot)
```

---

## Business Context

Telecom companies routinely lose 25–30% of their customer base annually. 
Every churned customer represents lost recurring revenue plus the cost of 
acquiring a replacement — typically 5–7x more expensive than retention.

Two questions drive this analysis:

1. Which customers are most likely to leave — and how confident is the model?
2. What is the financial case for intervening, and what does intervention cost?

The dataset contains 6,499 customer records across 17 features — contract 
type, tenure, service configuration, billing method, and churn status.

---

## Key Results

Before the technical detail, here is what the analysis produced.

**Model performance (Logistic Regression — best model):**

- AUC: 0.8355
- Accuracy at optimized threshold (0.6): 79.57%

**Applied to 500 new customers:**

| Metric | Value |
|--------|-------|
| Customers flagged as churn risk | 74 (14.8%) |
| Monthly revenue at risk | $6,158 |
| Proposed incentive cost | $2,220 ($30 per churner) |
| Net benefit if churners retained | $3,938 |
| ROI on incentive spend | 177% |

**Loyalty program extension (internet customers only):**

| Metric | Value |
|--------|-------|
| Eligible at-risk customers | 74 |
| Cost per customer | $259.72 |
| Net benefit per customer | $1,737.48 |
| Total retained revenue | $147,793 |
| Total net benefit | $128,574 |

---

## Churn Drivers

Five factors emerged as the strongest predictors from logistic regression 
odds ratios:

| Factor | Direction | Odds Ratio |
|--------|-----------|------------|
| Fiber optic internet | Increases churn | 3.03x |
| Month-to-month contract | Increases churn | Reference (highest risk) |
| Electronic check payment | Increases churn | 1.42x |
| Senior citizen | Increases churn | 1.36x |
| Two-year contract | Decreases churn | 0.195x (80% less likely) |
| Tenure (each month) | Decreases churn | 0.938x |
| Online security | Decreases churn | 0.686x |

**The single most actionable lever: contract type.** 
Moving a month-to-month customer to a two-year contract 
reduces their churn probability by 80%.

---

## Retention Strategy

Three interventions ranked by expected ROI:

**1. Contract conversion campaign**
Target month-to-month customers with discounted annual contract offers. 
It's the highest-volume segment and the strongest predictor — best place to start.

**2. Early-tenure outreach**
Customers in their first 12 months churn at disproportionately high rates. 
Most early churners have already mentally left before anyone reaches out. Getting ahead of that window is the intervention.

**3. Fiber optic service review**
Fiber users are 3x more likely to churn despite paying premium prices. 
Paying premium prices and leaving at twice the rate of DSL users, something is off, either on pricing or service delivery.

---

## Data Preparation

#### 1. Load and Inspect Data

```{r}
load("Model_Building_Data.rda")
str(Model_Building_Data)
summary(Model_Building_Data)
colSums(is.na(Model_Building_Data))
```

**Note:** 6,499 records, 17 features. TotalCharges has 11 missing values 
from blank entries stored as character — handled below.

#### 2. Clean and Type-Convert

```{r}
# TotalCharges stored as character with blanks — convert and remove NAs
Model_Building_Data$TotalCharges <- as.numeric(as.character(Model_Building_Data$TotalCharges))
Model_Building_Data <- na.omit(Model_Building_Data)

# Convert character columns and SeniorCitizen to factors
Model_Building_Data <- Model_Building_Data %>%
  mutate(across(where(is.character), as.factor),
         SeniorCitizen = as.factor(SeniorCitizen))
```

11 rows removed. Final dataset: 6,488 records, class balance maintained 
(~73% retained, ~27% churned).

---

## Exploratory Analysis

#### Churn Distribution

```{r}
ggplot(Model_Building_Data, aes(Status, fill = Status)) +
  geom_bar() +
  scale_fill_manual(values = c("Current" = "#3B82F6", "Left" = "#EF4444")) +
  labs(title = "Customer Churn Distribution", x = NULL, y = "Count") +
  theme_minimal() +
  theme(legend.position = "none")
```

The dataset is imbalanced — ~73% retained, ~27% churned. 
A naive model that always predicts 'retained' would already hit 73% accuracy — which is why threshold tuning matters here.

#### Tenure vs Churn

```{r}
ggplot(Model_Building_Data, aes(x = Status, y = Tenure, fill = Status)) +
  geom_boxplot() +
  scale_fill_manual(values = c("Current" = "#3B82F6", "Left" = "#EF4444")) +
  labs(title = "Tenure Distribution by Churn Status",
       x = NULL, y = "Tenure (months)") +
  theme_minimal() +
  theme(legend.position = "none")
```

Churned customers have median tenure well below retained customers. 
Month 1–12 is where most of the churn happens.

#### Monthly Charges vs Churn

```{r}
ggplot(Model_Building_Data, aes(x = Status, y = MonthlyCharges, fill = Status)) +
  geom_boxplot() +
  scale_fill_manual(values = c("Current" = "#3B82F6", "Left" = "#EF4444")) +
  labs(title = "Monthly Charges by Churn Status",
       x = NULL, y = "Monthly Charges ($)") +
  theme_minimal() +
  theme(legend.position = "none")
```

Churned customers pay higher monthly charges on average — 
paying more per month and leaving sooner go together.

#### Contract Type

```{r}
ggplot(Model_Building_Data, aes(x = Contract, fill = Status)) +
  geom_bar(position = "fill") +
  scale_y_continuous(labels = scales::percent) +
  scale_fill_manual(values = c("Current" = "#3B82F6", "Left" = "#EF4444")) +
  labs(title = "Churn Rate by Contract Type",
       x = NULL, y = "Proportion") +
  theme_minimal()
```

Month-to-month customers churn at dramatically higher rates. 
Two-year contract holders barely churn at all.

#### Internet Service

```{r}
ggplot(Model_Building_Data, aes(x = InternetService, fill = Status)) +
  geom_bar(position = "fill") +
  scale_y_continuous(labels = scales::percent) +
  scale_fill_manual(values = c("Current" = "#3B82F6", "Left" = "#EF4444")) +
  labs(title = "Churn Rate by Internet Service Type",
       x = NULL, y = "Proportion") +
  theme_minimal()
```

Fiber optic customers churn at nearly double the rate of DSL customers — 
despite paying more. That 3.03x odds ratio is the single largest in the model.

#### Payment Method

```{r}
ggplot(Model_Building_Data, aes(x = PaymentMethod, fill = Status)) +
  geom_bar(position = "fill") +
  scale_y_continuous(labels = scales::percent) +
  scale_fill_manual(values = c("Current" = "#3B82F6", "Left" = "#EF4444")) +
  labs(title = "Churn Rate by Payment Method",
       x = NULL, y = "Proportion") +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 30, hjust = 1))
```

Electronic check users churn at notably higher rates. 
Customers on auto-pay — bank transfer or credit card — churn at noticeably lower rates.

---

## Train-Test Split

```{r}
set.seed(123)
split <- createDataPartition(Model_Building_Data$Status, p = 0.8, list = FALSE)
train_data <- Model_Building_Data[split, ]
test_data  <- Model_Building_Data[-split, ]

prop.table(table(train_data$Status))
prop.table(table(test_data$Status))
```

80/20 stratified split. Class balance preserved in both subsets 
(~73% Current / ~27% Left in each).

---

## Model Building

Five models were built and evaluated on AUC and threshold-optimized accuracy. 
Each model uses the same train/test split for fair comparison.

### 5.1 Logistic Regression

```{r}
logit_model_full <- glm(Status ~ ., data = train_data, family = binomial)
summary(logit_model_full)
```

```{r}
test_probs <- predict(logit_model_full, newdata = test_data, type = "response")

roc_logit <- roc(test_data$Status, test_probs)
plot(roc_logit, col = "#2563EB", lwd = 2,
     main = "ROC Curve — Logistic Regression")
auc(roc_logit)
```

```{r}
# Threshold tuning
thresholds <- seq(0.1, 0.9, by = 0.1)
accuracy <- sapply(thresholds, function(th) {
  mean(ifelse(test_probs > th, "Left", "Current") == test_data$Status)
})

plot(thresholds, accuracy, type = "l", col = "#2563EB", lwd = 2,
     xlab = "Threshold", ylab = "Accuracy",
     main = "Threshold Tuning — Logistic Regression")

best_thresh <- thresholds[which.max(accuracy)]
cat("Best threshold:", best_thresh, "\n")
cat("Best accuracy:", max(accuracy), "\n")
```

```{r}
final_preds <- ifelse(test_probs > best_thresh, "Left", "Current")
confusionMatrix(as.factor(final_preds), test_data$Status, positive = "Left")
```

```{r}
# Odds ratios
odds_ratios <- sort(exp(coef(logit_model_full)), decreasing = TRUE)
odds_ratios
```

AUC: **0.8355** | Best accuracy: **79.57%** at threshold 0.6

Fiber optic (3.03x), electronic check (1.42x), and senior citizen status (1.36x) 
are the strongest positive churn predictors. 
Two-year contracts (0.195x) are the strongest retention signal.

---

### 5.2 Naive Bayes

```{r}
nb_model <- naive_bayes(Status ~ ., data = train_data)
nb_probs <- predict(nb_model, newdata = test_data, type = "prob")[, "Left"]

nb_roc <- roc(test_data$Status, nb_probs)
plot(nb_roc, col = "#7C3AED", lwd = 2,
     main = "ROC Curve — Naive Bayes")
auc(nb_roc)
```

```{r}
nb_accuracy <- sapply(thresholds, function(th) {
  mean(ifelse(nb_probs > th, "Left", "Current") == test_data$Status)
})

plot(thresholds, nb_accuracy, type = "l", col = "#7C3AED", lwd = 2,
     xlab = "Threshold", ylab = "Accuracy",
     main = "Threshold Tuning — Naive Bayes")

nb_best_thresh <- thresholds[which.max(nb_accuracy)]
cat("Best threshold:", nb_best_thresh, "\n")
```

```{r}
nb_final_preds <- ifelse(nb_probs > nb_best_thresh, "Left", "Current")
confusionMatrix(as.factor(nb_final_preds), test_data$Status, positive = "Left")
```

AUC: **0.8112** | Best accuracy: **77.10%** at threshold 0.9

---

### 5.3 Linear Discriminant Analysis (LDA)

```{r}
lda_model <- lda(Status ~ ., data = train_data)
lda_pred  <- predict(lda_model, newdata = test_data)
lda_probs <- lda_pred$posterior[, "Left"]

lda_roc <- roc(test_data$Status, lda_probs)
plot(lda_roc, col = "#F59E0B", lwd = 2,
     main = "ROC Curve — LDA")
auc(lda_roc)
```

```{r}
lda_accuracy <- sapply(thresholds, function(th) {
  mean(ifelse(lda_probs > th, "Left", "Current") == test_data$Status)
})

plot(thresholds, lda_accuracy, type = "l", col = "#F59E0B", lwd = 2,
     xlab = "Threshold", ylab = "Accuracy",
     main = "Threshold Tuning — LDA")

lda_best_thresh <- thresholds[which.max(lda_accuracy)]
cat("Best threshold:", lda_best_thresh, "\n")
```

```{r}
lda_final_preds <- ifelse(lda_probs > lda_best_thresh, "Left", "Current")
confusionMatrix(as.factor(lda_final_preds), test_data$Status, positive = "Left")
```

AUC: **0.8289** | Best accuracy: **79.11%** at threshold 0.6

---

### 5.4 Quadratic Discriminant Analysis (QDA)

```{r}
# Collapse "No internet/phone service" levels for QDA compatibility
qda_train <- train_data
qda_test  <- test_data

cols_to_clean <- c("MultipleLines", "OnlineSecurity", "DeviceProtection", "StreamingMovies")
for (col in cols_to_clean) {
  for (df in c("qda_train", "qda_test")) {
    temp <- get(df)
    temp[[col]] <- as.character(temp[[col]])
    temp[[col]][temp[[col]] %in% c("No internet service", "No phone service")] <- "No"
    temp[[col]] <- as.factor(temp[[col]])
    assign(df, temp)
  }
}

qda_model <- qda(Status ~ ., data = qda_train)
qda_pred  <- predict(qda_model, newdata = qda_test)
qda_probs <- qda_pred$posterior[, "Left"]

qda_roc <- roc(qda_test$Status, qda_probs)
plot(qda_roc, col = "#EF4444", lwd = 2,
     main = "ROC Curve — QDA")
auc(qda_roc)
```

```{r}
qda_accuracy <- sapply(thresholds, function(th) {
  mean(ifelse(qda_probs > th, "Left", "Current") == qda_test$Status)
})

plot(thresholds, qda_accuracy, type = "l", col = "#EF4444", lwd = 2,
     xlab = "Threshold", ylab = "Accuracy",
     main = "Threshold Tuning — QDA")

qda_best_thresh <- thresholds[which.max(qda_accuracy)]
cat("Best threshold:", qda_best_thresh, "\n")
```

```{r}
qda_final_preds <- ifelse(qda_probs > qda_best_thresh, "Left", "Current")
confusionMatrix(as.factor(qda_final_preds), qda_test$Status, positive = "Left")
```

AUC: **0.8180** | Best accuracy: **78.10%** at threshold 0.9

---

### 5.5 Decision Tree (Pruned)

```{r}
tree_model <- rpart(Status ~ ., data = train_data, method = "class", cp = 0.001)
printcp(tree_model)

opt_cp <- tree_model$cptable[which.min(tree_model$cptable[, "xerror"]), "CP"]
cat("Optimal cp:", opt_cp, "\n")
```

```{r}
pruned_tree <- prune(tree_model, cp = opt_cp)

par(mfrow = c(1, 2))
rpart.plot(tree_model,  main = "Before Pruning", extra = 106)
rpart.plot(pruned_tree, main = "After Pruning",  extra = 106)
par(mfrow = c(1, 1))
```

```{r}
tree_probs <- predict(pruned_tree, newdata = test_data, type = "prob")[, "Left"]

tree_roc <- roc(test_data$Status, tree_probs)
plot(tree_roc, col = "#10B981", lwd = 2,
     main = "ROC Curve — Pruned Decision Tree")
auc(tree_roc)
```

```{r}
tree_accuracy <- sapply(thresholds, function(th) {
  mean(ifelse(tree_probs > th, "Left", "Current") == test_data$Status)
})

plot(thresholds, tree_accuracy, type = "l", col = "#10B981", lwd = 2,
     xlab = "Threshold", ylab = "Accuracy",
     main = "Threshold Tuning — Decision Tree")

tree_best_thresh <- thresholds[which.max(tree_accuracy)]
cat("Best threshold:", tree_best_thresh, "\n")
```

```{r}
tree_final_preds <- ifelse(tree_probs > tree_best_thresh, "Left", "Current")
confusionMatrix(as.factor(tree_final_preds), test_data$Status, positive = "Left")
```

AUC: **0.8006** | Best accuracy: **78.95%** at threshold 0.5

Pruning reduced the tree from a complex overfit structure to a 
clean, interpretable model using Contract, InternetService, 
Tenure, and PaymentMethod as the primary splits.

---

## Model Comparison

```{r}
auc_logit <- auc(roc(test_data$Status, test_probs))
auc_nb    <- auc(roc(test_data$Status, nb_probs))
auc_lda   <- auc(roc(test_data$Status, lda_probs))
auc_qda   <- auc(roc(qda_test$Status, qda_probs))
auc_tree  <- auc(roc(test_data$Status, tree_probs))

acc_logit <- max(sapply(thresholds, function(th) mean(ifelse(test_probs  > th, "Left", "Current") == test_data$Status)))
acc_nb    <- max(sapply(thresholds, function(th) mean(ifelse(nb_probs    > th, "Left", "Current") == test_data$Status)))
acc_lda   <- max(sapply(thresholds, function(th) mean(ifelse(lda_probs   > th, "Left", "Current") == test_data$Status)))
acc_qda   <- max(sapply(thresholds, function(th) mean(ifelse(qda_probs   > th, "Left", "Current") == qda_test$Status)))
acc_tree  <- max(sapply(thresholds, function(th) mean(ifelse(tree_probs  > th, "Left", "Current") == test_data$Status)))

comparison <- data.frame(
  Model         = c("Logistic Regression", "LDA", "QDA", "Decision Tree", "Naive Bayes"),
  AUC           = round(c(auc_logit, auc_lda, auc_qda, auc_tree, auc_nb), 4),
  Best_Accuracy = round(c(acc_logit, acc_lda, acc_qda, acc_tree, acc_nb), 4),
  Best_Threshold = c(0.6, 0.6, 0.9, 0.5, 0.9)
)

knitr::kable(comparison,
             caption = "Model Comparison — AUC and Threshold-Optimized Accuracy",
             col.names = c("Model", "AUC", "Best Accuracy", "Best Threshold"))
```

**Logistic regression is the recommended model.** It leads on both AUC 
and accuracy, and its coefficients are directly interpretable as odds ratios — 
and its coefficients read directly as odds ratios, which gives a retention team something concrete to act on.

---

## Business Application

The logistic regression model was applied to a held-out implementation 
dataset of 500 customers to simulate real deployment.

#### Load and Prepare Implementation Data

```{r}
load("Implementation_Data.rda")

Implementation_Data <- Implementation_Data %>%
  mutate(across(where(is.character), as.factor),
         SeniorCitizen = as.factor(SeniorCitizen))

Implementation_Data$TotalCharges <- as.numeric(as.character(Implementation_Data$TotalCharges))
Implementation_Data <- na.omit(Implementation_Data)
```

#### Score Customers

```{r}
impl_probs <- predict(logit_model_full, newdata = Implementation_Data, type = "response")
impl_preds <- ifelse(impl_probs > best_thresh, "Left", "Current")

Implementation_Data$PredictedStatus    <- impl_preds
Implementation_Data$ChurnProbability   <- round(impl_probs, 4)
```

#### Revenue at Risk

```{r}
num_churn_risk  <- sum(Implementation_Data$PredictedStatus == "Left")
revenue_at_risk <- sum(Implementation_Data$MonthlyCharges[Implementation_Data$PredictedStatus == "Left"])

cat("Customers flagged as churn risk:", num_churn_risk, "\n")
cat("Monthly revenue at risk: $", round(revenue_at_risk, 2), "\n")
```

#### Incentive Simulation

```{r}
discount_per_month <- 10
discount_months    <- 3
cost_per_churner   <- discount_per_month * discount_months

total_incentive_cost <- num_churn_risk * cost_per_churner
net_benefit          <- revenue_at_risk - total_incentive_cost
roi                  <- (net_benefit / total_incentive_cost) * 100

cat("Incentive cost per customer: $", cost_per_churner, "\n")
cat("Total incentive spend: $", total_incentive_cost, "\n")
cat("Revenue retained if successful: $", round(revenue_at_risk, 2), "\n")
cat("Net benefit: $", round(net_benefit, 2), "\n")
cat("ROI on incentive spend:", round(roi, 1), "%\n")
```

#### Loyalty Program (Internet Customers)

```{r}
churn_internet <- Implementation_Data %>%
  filter(PredictedStatus == "Left", InternetService != "No")

avg_monthly_bill    <- mean(churn_internet$MonthlyCharges)
discount_24_months  <- 0.10 * avg_monthly_bill * 24
addon_cost          <- 5 * 12
cost_per_customer   <- discount_24_months + addon_cost
retained_per_customer <- avg_monthly_bill * 24
benefit_per_customer  <- retained_per_customer - cost_per_customer

n_eligible            <- nrow(churn_internet)
total_program_cost    <- cost_per_customer * n_eligible
total_retained        <- retained_per_customer * n_eligible
total_net_benefit     <- benefit_per_customer * n_eligible

cat("Eligible customers:", n_eligible, "\n")
cat("Avg monthly bill: $", round(avg_monthly_bill, 2), "\n")
cat("Cost per customer: $", round(cost_per_customer, 2), "\n")
cat("Net benefit per customer: $", round(benefit_per_customer, 2), "\n")
cat("Total program cost: $", round(total_program_cost, 2), "\n")
cat("Total retained revenue: $", round(total_retained, 2), "\n")
cat("Total net benefit: $", round(total_net_benefit, 2), "\n")
```

---

## Summary

**Best model:** Logistic Regression — AUC 0.8355, accuracy 79.57% at threshold 0.6

**Top churn drivers:**
- Fiber optic internet (3.03x more likely to churn vs DSL)
- Month-to-month contract (reference group — highest risk)
- Electronic check payment (1.42x)
- Two-year contract (0.195x — strongest retention signal)

**Financial case for action (500 customers):**

| Intervention | Cost | Revenue Protected | Net Benefit |
|---|---|---|---|
| Short incentive ($30/churner) | $2,220 | $6,158 | $3,938 |
| Loyalty program (internet users) | $19,219 | $147,793 | $128,574 |

Any CRM that takes a probability score can use this directly — flag customers above 0.6, trigger the outreach. The short incentive returns $1.77 for every dollar spent. The loyalty program returns $6.69.
